UMAP reveals cryptic population structure and phenotype heterogeneity in large genomic cohorts

### 2019-07-05 ms/main.tex
more FEV1 explanation, add some citations to intro, clarify colour scheme in S4

%An Analysis is a new analysis of existing data (typically large genomic, transcriptomic or proteomic data sets from arrays or other high-throughput platforms) or new data obtained in a comparative analysis of technologies that lead to novel and arresting conclusions of importance to a broad audience. The main text (excluding abstract, Methods, references and figure legends) is approximately 3,000 words. The abstract is typically 100-150 words, unreferenced. Analyses have no more than 6 display items (figures and/or tables). An introduction (without heading) is followed by sections headed Results, Discussion and online Methods. The Results and online Methods should be divided by topical subheadings; the Discussion does not contain subheadings. As a guideline, Analysis allow up to 50 references.

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference} \sgcomment{This is a lone reference for a broad topic... Are there more recent reviews perhaps? Else add a couple of recent references eg by the novembre lab?}.

@@

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either method on the top principal components leads to distinct population clusters and less defined continental structure. Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}). To investigate the population information contained in low-variance PCs, we performed UMAP on data projected onto PCs 100 to 3450 (i.e., without information about the leading 99 PCs). Figure~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible. \sgcomment{Note figure S4 lacks a color code -- at least point out that it uses the same coloring as X}

@@

\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage.}
Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan. Baseline FEV1 levels are determined via equations that include ancestry or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}; we are not aware of any results comparing populations within Asia. We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines (the largest populations in the UKBB by country of birth) as well as the self-identified Chinese population (mutually exclusive groups, shown in figure~\ref{fig:supp_fev_test_pops}). After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$). By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}). For comparison, the adjusted difference between self-identified African and British individuals in the UKBB is 0.762 (p-value $2.2\times 10^{-16}$).

%For reference, \sgcomment{US organisation} recommends using different FEV1 reference scales for asthma diagnostic in European-Americans and African-Americans based on a population mean difference in FEV1 of XX.

%Some of the phenotypic variation patterns mentioned above could, in principle, have been identified by considering large numbers of box plots, but we find that UMAP facilitates the identification of large phenotypic differences among genetically related groups. This is especially useful when population labels based on self-reported identity or country of birth are unavailable or correlate poorly with genetic ancestry.

\subsubsection*{Comparing t-SNE and UMAP}
\sgcomment{tweaked this a bit -- would be good to include in reply to reviewer}
Identifying the best dimension reduction technique is challenging, both because the `best' representation depends on context, and because convergence issues may mean that a good theoretical model for dimensional reduction might perform poorly because of challenges in numerical optimization. To assess whether the relatively poor performance of t-SNE could be due to convergence rather than a flawed model, we used UMAP to preprocess the UKBB data and provide a starting point to a standard t-SNE implementation. This led to representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).  Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{images/1KGP_UMAP_PCS100_PCE3450_NC2_NN15_MD05_201944185843.jpeg}
    \end{subfigure}
    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation.}
    \label{fig:supp_1kgp_3350}
\end{figure*}

>>>>

@@

\maketitle
\thispagestyle{firststyle}
\ifthenelse{\boolean{shortarticle}}{\ifthenelse{\boolean{singlecolumn}}{\abscontentformatted}{\abscontent}}{}
Questions in medicine, anthropology, and related fields hinge on interpreting the deluge of genomic data provided by modern high-throughput sequencing technologies. Because genomic datasets are high-dimensional, their interpretation requires statistical methods that can comprehensively condense information in a manner that is understandable to researchers and minimizes the amount of data that is sacrificed. Both model-based and model-agnostic approaches to summarize data have played important roles in shaping our understanding of the evolution of our species~\cite{lawson2012inference, novembre2016recent, spence2018inference}.\adpcomment{added two citations}

@@

\subsection*{Results}
\subsubsection*{Fine-scale visualization of the 1KGP dataset} The 1KGP contains genotype data of 3,450 individuals from 26 relatively distinct labeled populations\cite{10002015global}. Figure~\ref{fig:1000g_compare} shows visualizations using PCA, t-SNE, UMAP, and UMAP with PCA pre-processing. Using UMAP and t-SNE on the genotype data presents clusters that are roughly grouped by continent, with UMAP showing a clear hierarchy of population and continental clusters, whereas t-SNE fails to assign many individuals to population clusters. Using either method on the top principal components leads to distinct population clusters and less defined continental structure. Adding more components results in progressively finer clusters until approximately 20 populations appear using 15 components; adding further components converges to results similar to using the entire genotype data (see figures~\ref{fig:supp_megamontage_pc2_9}, \ref{fig:supp_megamontage_pc10_50}, and \ref{fig:supp_montage_1kgp_converge}). To investigate the population information contained in low-variance PCs, we performed UMAP on data projected onto PCs 100 to 3450 (i.e., without information about the leading 99 PCs). Figure~\ref{fig:supp_1kgp_3350} shows that population structure is still clearly visible.

@@

\adpcomment{Discussing FEV1 here. Note: citations tend to present relative differences b/w groups as a percentage so I used UKBB data.}
Figure~\ref{fig:supp_comparison_fev_chi_eur} further suggests a difference in FEV1 between those who self-identified as Chinese and a nearby cluster enriched in individuals born in Japan; to our knowledge there have not been studies into differences in FEV1 between these populations. Baseline FEV1 levels are determined via equations that include ethnicity or race\cite{quanjer2012multi}, but studies in several populations have shown that there is considerable variation based on ancestry, even within self-defined ethnicity\cite{ortega2015effect}. We selected the individuals whose UMAP coordinates were near the Chinese cluster and kept those born in Japan, Malaysia, and the Philippines as well as the self-identified Chinese population. These four groups are mutually exclusive and are shown in figure~\ref{fig:supp_fev_test_pops}. Using this approach facilitates identifying large phenotypic differences among genetically related groups, which is useful when population labels are unavailable or correlate poorly with genetic ancestry. After adjusting for age, age$^2$, height, and sex, an unpaired two-sample t-test shows those born in Japan have a higher mean FEV1 than Chinese individuals by 0.224 (p-value $2.787\times 10^{-15}$). By sex, there is a difference of 0.213 (p-value $5.4\times 10^{-13}$) among females and 0.317 (p-value $5.1\times 10^{-4}$) among males, though there are considerably fewer males in the sample (distributions presented in figure~\ref{fig:supp_fev_ridgeplots}). For comparison, the adjusted difference between self-identified African and British individuals in the UKBB is 0.762 (p-value $2.2\times 10^{-16}$).

\subsubsection*{Comparing t-SNE and UMAP}
%\sgcomment{tweaked this a bit -- would be good to include in reply to reviewer}
Identifying the best dimension reduction technique is challenging, both because the `best' representation depends on context, and because convergence issues may mean that a good theoretical model for dimensional reduction might perform poorly because of challenges in numerical optimization. To assess whether the relatively poor performance of t-SNE could be due to convergence rather than a flawed model, we used UMAP to preprocess the UKBB data and provide a starting point to a standard t-SNE implementation. This led to representations that were objectively better (according to the t-SNE metric) than the default t-SNE implementation (figure~\ref{fig:supp_tsne_umap_compare_ukbb_graph}). Yet, these representations were much less detailed than the UMAP embedding provided as a starting point (figure~\ref{fig:supp_tsne_umap_compare_ukbb}).  Given these results, we recommend UMAP over t-SNE for large and diverse genomic datasets.

@@

\begin{figure*}
    \centering
    \begin{subfigure}{0.95\textwidth}
    \includegraphics[width=0.95\textwidth]{images/1KGP_UMAP_PCS100_PCE3450_NC2_NN15_MD05_201944185843.jpeg}
    \end{subfigure}
    \caption{UMAP applied the last 3350 principal components of the 1KGP, which explain 78.7\% of the variation. The colour scheme is the same as in figure~\ref{fig:subpops}.}
    \label{fig:supp_1kgp_3350}
\end{figure*}
